---
title: "R_for_Data_Science_ASSESSMENT"
author: "B270551"
date: "2024-11-28"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align="center")

```

# B270551 - RDS Class Assignment:

Creating a custom workflow for processing and visualising data 

```{r header, include=FALSE}

```

## First, we need to load in the relevant data sets:  

I have named these 'annot', 'data' and 'genelist' for gene annotation, full data & the relevant gene subset respectively.
```{r loading_data, include=TRUE}
annot <- read.csv("./gene_annotation.csv")
data <- read.csv("./data_all.csv")
genelist <- read.csv("./genelist_6.txt")
```

## Now, loading in the relevant libraries:  


```{r loading_packages, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
install.packages("tidyverse")
library(tidyverse)
library(pheatmap)
```

## Genelist Manipulation 

Our genelist contains duplicates, which must first be removed: 
```{r manipulating data, include=TRUE}
#Made a boolean vector of duplicates and used this to index the genelist
duplicates <- duplicated(genelist$x)
genelist <- genelist[!duplicates,]
#...and now need to restore list to a data frame
genelist <- as.data.frame(genelist)
```

## Dataset Manipulation

In this stage, I am manipulating my dataset. I have **logged all values**,  **replaced gene index with the gene longname** and have **annotated each gene with its 'type'**.

I have then indexed the data with the unique genelist from above to extract out only the relevant genes.

The first few rows of the output is as follows:

```{r datamanipulation, echo=FALSE}
data <- log(data)
data$longname <- annot$LongName
data$type <- as.character(annot$Type)
#removing 'X' column of gene numbers
data <- data[,2:15]
#moving longname and type columns to left hand side
data <- data[, c((ncol(data)-1):ncol(data), 1:(ncol(data)-2))]
#Selecting the appropriate data by indexing GeneList
gene_subset <- genelist [, "genelist"]
relevant_data <- data[gene_subset,]
head(relevant_data)
```
This is our base data frame we can now use for our plots.
<b>

## Generating a Boxplot 

### i) Modifying our base data 

To make our base data suitable for a boxplot, we can generate a pivot table for including the treatment group.

The output is as follows:
```{r pivot_table, echo=FALSE}
#Creating a pivot table of relevant data
pivot_relevant_data <- relevant_data %>%
  pivot_longer(cols = A:L,
               names_to = "variable",
               values_to = "values")

#Adding Treatments to pivot table data 
pivot_relevant_data$Treatment <- as.character(rep(c(1, 1, 1, 2, 2, 2, 3, 3, 3, 4, 4, 4), times = 40))
#Ensuring type is as.character
pivot_relevant_data$type <- as.character(pivot_relevant_data$type)

head(pivot_relevant_data)
```
### ii) Generating the plot 

We can now use this data to generate our box plot, where jitter is applied to separate out data for each genetype within each treatment group:
```{r plotting boxplot, echo=FALSE, fig.align='center'}
treatment_colours <- c("1" = "lightblue", "2" = "skyblue", "3" = "deepskyblue", "4" = "dodgerblue")
  point_colours <- setNames(c("slateblue3", "lavender", "blueviolet"), c("XA", "XB", "XC")) 
 
  boxplot(
    values ~ Treatment, 
    data = pivot_relevant_data,
    col = treatment_colours,           # Colors for each treatment
    border = "black",           # Border color for boxes
    xlab = "Treatment Group",   # X-axis label
    ylab = "Log(Expression)",   # Y-axis label
    main = "Figure 1: Log(Expression) for Genes in each Treatment Group",  
    notch = FALSE,              
    las = 1                     # Using to rotate axis labels for readability
  )
  
  # Overlay jittered points (XA)
  stripchart(
    values ~ (Treatment), #shifting so on LHS
    data = subset(pivot_relevant_data, type == "XA"),
    method = "jitter",          # Add jitter to reduce overlap
    vertical = TRUE,            # Arrange points vertically
    pch = 19,
    col =  "slateblue3",
    add = TRUE,                 # Add to the existing boxplot
    cex = 0.8,                  # Size of the points
    jitter = 0.2,                # Amount of horizontal jitter
    at = as.numeric(unique(pivot_relevant_data$Treatment)) - 0.2
  )
  
  # Overlay jittered points (XB)
  stripchart(
    values ~ Treatment,
    data = subset(pivot_relevant_data, type == "XB"),
    method = "jitter",          # Add jitter to reduce overlap
    vertical = TRUE,            # Arrange points vertically
    pch = 19,
    col =  "lavender",
    add = TRUE,                 # Add to the existing boxplot
    cex = 0.8,                  # Size of the points
    jitter = 0.2                # Amount of horizontal jitter
  )
  
  # Overlay jittered points (XC)
  stripchart(
    values ~ (Treatment), #shifting so on RHS
    data = subset(pivot_relevant_data, type == "XC"),
    method = "jitter",          # Add jitter to reduce overlap
    vertical = TRUE,            # Arrange points vertically
    pch = 19,
    col =  "blueviolet",
    add = TRUE,                 # Add to the existing boxplot
    cex = 0.8,                  # Size of the points
    jitter = 0.2,                # Amount of horizontal jitter
    at = as.numeric(unique(pivot_relevant_data$Treatment)) + 0.2,
  )
  
  # Adding a legend
  legend(
    "topright", 
    legend = unique(pivot_relevant_data$type), 
    pch = 19, 
    col = point_colours, 
    title = "Type"
  )
```

## Generating heatmaps 

### i) Manipulating Data 

We need to manipulate our base data in a slightly different way for the heatmaps to ensure rownames are relabelled as the gene long name

We also need to ensure relevant datasets exist that can be used to generate the row and column annotations:
```{r heatmap manipulation, include=TRUE}
#Creating Heatmaps 
heatmap_data <- relevant_data
rownames(heatmap_data) <- heatmap_data$longname
heatmap_data <- heatmap_data[,3:14]
heatmap_data <- as.matrix(heatmap_data) 
write.csv(heatmap_data, file= "heatmap_data.csv")  

#Creating data for annotations (rows)
gene_types <- data.frame(Type = relevant_data$type)
rownames(gene_types) <- relevant_data$longname 


#Creating data dor annotations (columns)
treatment <- data.frame(
  Treatment = rep(c("1", "2", "3", "4"), each = 3)  # Treatments
)
rownames(treatment) <- colnames(heatmap_data)

#Let's view our heatmap data to chcek it is as we want:
head(heatmap_data)
```
### ii) Generating the heatmaps 

First, generating a heatmap where only the genes are clustered:

```{r heatmap1, echo=FALSE, fig.align='center'}
annotation_colours <- list(
  Type = c(XA = "slateblue3", XB = "lavender", XC = "blueviolet"),
  Treatment = c("1" = "SlateGray", "2" = "Navy", "3" = "SteelBlue", "4" = "DodgerBlue4")
)

pheatmap(heatmap_data,
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         scale='row',
         fontsize_row = 6,
         fontsize_col = 10,
         legend=TRUE,
         main = "Figure 2 - Heatmap: Only Genes Clustered",
         annotation_row = gene_types,
         annotation_col = treatment,
         annotation_colors = annotation_colours,
         annotation_border = TRUE,
         color = colorRampPalette(c("navy", "white", "firebrick3"))(500)
)

```

Now, generating a heatmap where BOTH genes and treatments are clustered:

```{r heatmap2, echo=FALSE, fig.align='centre'}
pheatmap(heatmap_data,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         scale='row',
         fontsize_row = 6,
         fontsize_col = 10,
         legend=TRUE,
         main = "Figure 3 - Heatmap: Genes and Samples Clustered",
         annotation_row = gene_types,
         annotation_col = treatment,
         annotation_colors = annotation_colours,
         annotation_border = TRUE,
         color = colorRampPalette(c("navy", "white", "firebrick3"))(500)
)

```

